---
title: "External C++ Functionality in Stan"
author: 
- "Matthew Amato-Yarbrough, mba75@nau.edu"
date: January 2023
output:
  pdf_document:
    toc: yes
    fig_crop: no
  html_document:
    toc: true 
    toc_depth: 2 
    theme: readable  
    toc_float: true
---

```{r, echo=FALSE}
# Initialization code chunk

# Source (modified): https://stackoverflow.com/a/53052436
# Text alignment: https://www.overleaf.com/learn/latex/Text_alignment
# Left section header
left_section_header = function(input_text){
  if(knitr::is_latex_output())
    paste0("\\begin{raggedright}\n 
           \\section{", 
           input_text, 
           "} 
           \n\\end{raggedright}")
  else if(knitr::is_html_output())
    paste0("<left>\n <section>\n <h1>", input_text, 
           "</h1> \n</section> \n</left>")
}

left_subsection_header = function(input_text){
  if(knitr::is_latex_output())
    paste0("\\begin{raggedright}\n 
           \\subsection{", 
           input_text, 
           "} 
           \n\\end{raggedright}")
  else if(knitr::is_html_output())
    paste0("<left>\n <section>\n <h2>", input_text, 
           "</h2> \n</section> \n</left>")
}

left_subsubsection_header = function(input_text){
  if(knitr::is_latex_output())
    paste0("\\begin{raggedright}\n 
           \\subsubsection{", 
           input_text, 
           "} 
           \n\\end{raggedright}")
  else if(knitr::is_html_output())
    paste0("<left>\n <section>\n <h3>", input_text, 
           "</h3> \n</section> \n</left>")
}

# Source (modified): https://stackoverflow.com/a/55064070
# Note: Not used for now
new_page_break = function() {
  if(knitr::is_latex_output())
    return("\\newpage")
  else if(knitr::is_html_output())
    return('<div style="page-break-before: always;" />')
  
}

# Checkout The R Markdown Cheat Sheet:
# https://www.rstudio.com/blog/the-r-markdown-cheat-sheet/
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 5)
```

\newpage

`r left_section_header("Introduction")`

The aim of this walk-through is to provide [\textbf{Stan}](https://mc-stan.org/) users  
with the knowledge of incorporating external C++ code into their preexisting Stan models. 
The main benefit of doing so is the ability to create and use custom C++ functions 
within your Stan model. 

The first example of this functionality will use the Bernoulli example code provided 
by the Stan developers. The code for this example is found in the 
`/cmdstan/examples/bernoulli` directory of a 
[\textbf{CmdStan installation}](https://github.com/stan-dev/cmdstan/tree/develop/examples/bernoulli). 
The second example will be a binomial model using data from 

[\textbf{Linear Models with R}](https://julianfaraway.github.io/faraway/LMR/). 
The code for this section will be provided below. 

There exists other Stan interfaces that provide external C++ functionality as well. 
A notable example is [\textbf{RStan}](http://mc-stan.org/rstan/). However, the 
difficulty with using external C++ code is relatively the same in both CmdStan 
and RStan.The deciding factors in choosing a Stan interface in this case are: 

- How comfortable you are with each of the interfaces.
- It is easier to enter data and set hyperparameters within RStan. 
- Power users will find CmdStan easier since data entry and hyperparameters are set through the terminal. 
- CmdStan is typically more-up-to-date when it comes the Stan back-end. 

With any Stan interface, the user will be required to [\textbf{template}](https://cplusplus.com/doc/oldtutorial/templates/)
their C++ code to match the templating found in [\textbf{Stan's math library}](https://mc-stan.org/math/). 
More detail will be provided below regarding the process of templating C++ code
to match the templating found in Stan's math library. This requirement remains the 
biggest time commitment in this process. 

Additional information on using external C++ code within RStan can be found in this 
[\textbf{vinette}](https://mc-stan.org/rstan/articles/external.html). Using 
external C++ code within RStan will not be explored in this walk-through. 

`r left_subsection_header("Installation")`

The first step is download the version of CmdStan appropriate for your system. 
The available versions of CmdStan are listed under releases in the 
[\textbf{stan-dev/cmdstan}](https://github.com/stan-dev/cmdstan/releases) 
repository. For example: The release that was used for this walk-through is 
[\textbf{cmdstan-2.29.2.tar.gz}](https://github.com/stan-dev/cmdstan/releases/download/v2.29.2/cmdstan-2.29.2.tar.gz). 

Once the tar.gz is downloaded and unzipped into a directory of your choosing 
(the system's home directory was used for this example), then you open a terminal 
window and cd into the unzipped folder (e.g. `cd ~/cmdstan-2.29.2/`). Within your 
terminal window, run `make build` to complete the CmdStan installation process. 
If you wanted to delete CmdStan from your system, you can run `make clean-all` 
from the same directory as you called `make build`. 

More recent versions of CmdStan can be used. However, the external C++ 
functionality may differ with future releases of CmdStan 

The details of the system used for this walkthrough: 

- Operating System: Ubuntu 20.04 amd64
- CmdStan Version: v2.29.2
- Compiler/Toolkit: g++ 9.4.0 and make 4.2.1

`r left_subsection_header("Repository")`

For detailed terminal output and what you should expect to see after running the 
`make`, `sample`, and `stansummary` commands, refer to the "terminal_output" 
directories within the "bernoulli" and "binomial_glm" directories. These directories 
can be found in the same github repository associated with this documentation: 
"/external_cpp_functionality_in_stan". This repository also contains all of the 
code referred to within this walkthrough, as well as examples of CmdStan 
makefiles that can be used for each set of code. 

\newpage

`r left_section_header("Bernoulli example")`

This section will use the Bernoulli source code that is provided by the Stan developers. 
The associated Stan documentation that expands on the Bernoulli source code using 
external C++ code can be found [\textbf{here}](https://mc-stan.org/docs/cmdstan-guide/using-external-cpp-code.html). 
A condensed version of this documentation will be found below, but it is recommended 
that you read the Stan documentation if you want additional detail. 

First step: Within the Bernoulli directory of your CmdStan installation, open up 
the file titled "bernoulli.stan". If you installed CmdStan into your home directory, 
then this file would be found in ~/cmdstan-2.29.2/examples/bernouli. 

Above the `data` block, add the following lines of code: 

```
functions {
  real make_odds(real theta);
}
``` 

Below the `model` block, add the additional lines of code: 

```
generated quantities {
  real odds;
  odds = make_odds(theta);
}
```

This is the essential first step to editing your Stan code so that it can use 
external C++ code. You are adding a function called `make_odds` that accepts a 
`real` data type (`theta`) and outputs a `real` data type. Additional information 
on data types within Stan can be found [\textbf{here}](https://mc-stan.org/docs/reference-manual/overview-of-data-types.html). 
The `make_odds` function will be the function defined within your external C++ file. 

Within the same bernouli directory, create a new file titled "make_odds.hpp". 
In "make_odds.hpp", add the following lines of code: 

```
#include <stan/model/model_header.hpp>
#include <ostream>

namespace bernoulli_model_namespace {
template <typename T0__,
          stan::require_all_t<stan::is_stan_scalar<T0__>>* = nullptr>
  stan::promote_args_t<T0__>
  make_odds(const T0__& theta, std::ostream* pstream__) {
    return theta / (1 - theta);
  }
}
```

For now, save and close "make_odds.hpp". A similar file will be revisited later 
with additional details. 

Open up a terminal window within the CmdStan directory (~/cmdstan-2.29.2/) and execute 
the following line: 

```
make STANCFLAGS=--allow-undefined USER_HEADER=examples/bernoulli/make_odds.hpp \
  examples/bernoulli/bernoulli_makeodds 
```

In this make call, both the `STANCFLAGS=--allow-undefined` and `USER_HEADER=` 
flags are necessary for allowing undefined functions within your Stan code and 
providing a location to an external header file. As noted by the Stan 
documentation: "you could put STANCFLAGS and USER_HEADER into the make/local 
file instead of specifying them on the command-line." 
This provides users with the option of using a simplified make call: 

```
make examples/bernoulli/bernoulli_makeodds 
```

Assuming that your `make` call was successful, you should find a file titled 
"bernoulli_makeodds" within your /bernoulli directory. This is the compiled 
executable that is a a result of your previous `make` call. 
It will contain the original bernoulli.stan model code, as well as the 
custom external C++ code. 


Note: Refer to "external_cpp_functionality_in_stan/bernoulli/terminal_output/" 
on github for an example of a successful `make` call would look like. 

Once you have the compiled executable the remaining CmdStan steps are the same. 

The Stan developers provide two data files for this example: 
"bernoulli.data.json" and "bernoulli.data.R". Below are the steps one would 
follow to sampling using the .json data file. 

Sample using the provided example data: `./bernoulli_makeodds sample data file=bernoulli.data.json`

Display using the `stansummary` functionality: `~/cmdstan-2.29.2/bin/stansummary output.csv`

These steps are discussed in detail in this [\textbf{part of the CmdStan documentation}](https://mc-stan.org/docs/cmdstan-guide/mcmc-intro.html). 
Additional arguments will be used in the next section. 

\newpage

`r left_section_header("Beyond the Bernoulli example")`

This section will provide a more complicated example of using external C++ code  
within CmdStan. 

Within the examples directory of your CmdStan installation, you will need to create 
a new directory as well as several new files. These files will be similar to the 
files from the previous section. However, the contents of these files will be 
different. Navigate to ~/cmdstan-2.29.2/examples/ and create a new directory 
titled "binomial_glm". Afterwards, open up the newly created directory a create 
the three following files: 

1. binomial_glm.stan
2. external_binomial_glm.hpp
3. bin_glm.Data.list.dump

Note: Depending on your preference, it could be easier to use the 
[\textbf{touch}](https://itslinuxfoss.com/touch-command-ubuntu-22-04/) 
command within terminal to create these files. 

`r left_subsection_header("Binomial GLM stan model")`

Open up binomial_glm.stan with a text editor and paste the following contents into 
the file and then save the file. 

```
functions {
    real custom_inv_logit(real value);
}

data {
    int<lower=1> N;                       // #obs
    int<lower=1> p;                       // #params
    int<lower=1> n[N];                    // #trials
    int<lower=0> y[N];                    // #successes 
    matrix<lower=0, upper=1> [N,p] X;     // X matrix
    real bkprior[2];                      // bk (k>0) prior mean and sd 
}

parameters {
    vector[p] b;
}

model {
    vector[N] output;

    output = X * b;
    for (i in 1:N) {
        // [y | beta]
        y[i] ~ binomial(n[i], custom_inv_logit(output[i]));
    }

    // [beta] = [b0] * prod_k [bk] (k>1)
    b[2:11] ~ normal(bkprior[1], bkprior[2]);
    
    // [b0 flat improper implied]
}
```

Explanation: The data and parameters block is fairly common for Stan code. The 
interesting points with the above code are the functions and model blocks. 

The functions block: `custom_inv_logit` is the name of the function found in 
the external C++ file. Similar to the Bernoulii example, the output of the 
function call is a real number and the input is a real number call value. 

The model block: The matrix and vector multiplication is done (`output = X * b`) 
within the Stan code, rather than within the external C++ code. While this 
adds additional lines of Stan code, it's to avoid the hassle of matrix and vector 
multiplication within the external C++ code. Refer to the "Matrices and vertices" 
subsection below for more information. 

`r left_subsection_header("External binomial hpp code")`

Now open up external_binomial_glm.hpp and paste the following code. Afterwards, 
save the file. 

```
#include "external_binomial_glm.hpp"
#include "stan/math/prim/fun/exp.hpp"

namespace binomial_glm_model_namespace {
template <typename T0__, stan::require_stan_scalar_t<T0__>* = nullptr>
stan::promote_args_t<T0__>
custom_inv_logit(const T0__& value, std::ostream* pstream__)
{
	stan::math::exp_fun exponent_function;

	if (value > 0) {
		return 1.0 / (1.0 + exponent_function.fun(-value));
	}
	else {
		T0__ e = exponent_function.fun(value);
		return e / (1.0 + e);
	}
}
}
```

Explanation: There are a handful of points worth mentioning here. 

Inverse logit: This was modeled after Stan's 
[\textbf{inv logit struct}](https://mc-stan.org/math/structstan_1_1math_1_1inv__logit__fun.html) 
found within the Stan library. While this function could be directly used within your 
Stan model, it was used as the external function within external_binomial_glm.hpp
for demonstrative purposes. 

Exponential function: The exponential function comes from Stan's 
[\textbf{exp fun struct}](https://mc-stan.org/math/structstan_1_1math_1_1exp__fun.html). 
It's important to note here that you could not get away with using the exponential 
function from C++'s standard library, as it is not built to accept Stan math types. 
As such, the best option is to use Stan's exp_fun struct. Once the struct option 
is created (called `exponent_function` above), you can access it's exponential 
function to pass in your `value` parameter. 

Creating a new variable: Notice that the variable `e` is of type `T0__`; the same 
type as the input parameter `value`. This is necessary in order to store the 
contents of the exponential function. 

Initial template: When attempting to compile external C++ code with CmdStan, it 
will create a separate header file from yours that contains the templated header 
for your external file. This will happen whether compilation is successful or fails. 

- For example: When attempting to run a `make` call with your binomial_glm.stan 
file completed but not your external C++ file, you will encounter 
a compilation error. But, within the same directory as your .stan file, there 
should be a file called binomial_glm.hpp. Open up that file and you should see the 
appropriate template for your external file towards the top portion of 
binomial_glm.hpp. You will need to include that template (as well as the namespace 
calls) within your header file. 

`r left_subsection_header("External binomial RDump file")`

Now that both the Stan model and external C++ code are ready, all that is left 
is to run the sampler. However, this will require a data file. 

Open up bin_glm.Data.list.dump, paste the following lines of code, save and then 
exit. 

```
N <- 24
p <- 11
n <- 
c(56, 34, 8, 82, 50, 18, 16, 39, 21, 31, 52, 40, 87, 56,
20, 35, 23, 9, 45, 72, 49, 28, 36, 24)
y <- 
c(9, 19, 7, 5, 6, 12, 5, 21, 19, 7, 14, 25, 58, 45, 20, 11,
7, 7, 32, 61, 42, 12, 15, 17)
X <- 
structure(c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1,
1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0,
0, 0, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1,
1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 0, 0, 1, 1, 1, 0, 0, 0, 1, 1, 1,
0, 0, 0, 1, 1, 1, 0, 0, 0, 1, 1, 1, 0, 1, 0, 0, 1, 0, 0, 1, 0, 0,
1, 0, 0, 1, 0, 0, 1, 0, 0, 1, 0, 0, 1, 0, 0, 0, 1, 0, 0, 1, 0, 0,
1, 0, 0, 1, 0, 0, 1, 0, 0, 1, 0, 0, 1, 0, 0, 1, 0, 0, 0, 0, 0, 0,
0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0,
0, 0, 0, 1, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 1, 0, 0, 0,
0, 0, 0, 0, 0, 0, 1, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 1,
0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 1, 0, 0, 1, 0, 0,
1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 1, 0, 0,
1, 0, 0, 1),
.Dim = c(24, 11))
bkprior <- 
c(1.77635683940025e-15, 4.69923960063867)
```

Explanation: This data set was compiled from Julian Faraway's data on head injury 
data. In order to get the data into a readable format for Stan (.RDump), Dr. Barber 
used the `rstan::stan_rdump` function within R. 

For more information on creating RDump data files for use in Cmdstan, refer to: 
[\textbf{RDump Format for CmdStan}](https://mc-stan.org/docs/cmdstan-guide/rdump.html).

Although JSON files were not used in this walkthrough, they should work similarly 
to RDump files, albeit the contents might be structured differently and the 
file extensions will be different when entering the data via the command line. 
For more information on creating JSON data files for use in Cmdstan, refer to: 
[\textbf{JSON Format for CmdStan}](https://mc-stan.org/docs/cmdstan-guide/json.html).

`r left_subsection_header("Final steps")`

Now that all of the necessary files have been provided, the process remains 
relatively the same as before: `make`, sample, display. 

To make the executable, open up a terminal window within the CmdStan directory 
(~/cmdstan-2.29.2/) and execute the following line: 

```
make STANCFLAGS=--allow-undefined USER_HEADER=examples/binomial_glm/external_binomial_glm.hpp \
  examples/binomial_glm/binomial_glm
```

If compilation was successful, you should find a file titled "binomial_glm" 
within your /binomial_glm directory. 

Sample using the provided example data: `./binomial_glm sample data file=binGLM.Data.list.dump.R`

For a more interesting sampling routine, additional arguments can be provided to 
CmdStan and used within a for loop to produce multiple chains: 

```
for i in {1..3}; do ./binomial_glm sample random seed=110710 \
  data file=binGLM.Data.list.dump.R output refresh=1000 file=output_${i}.csv; done;
```

Display using the `stansummary` functionality: `~/cmdstan-2.29.2/bin/stansummary output_*.csv`


\newpage

`r left_section_header("Additional points")`

`r left_subsection_header("Adding more inputs")`

Going off of the "Beyond the Bernoulli example" section above, you will  
likely need to add more than one input to the your external C++ code to get the
most out of it. Assuming that the data type that you want to add is covered by 
Stan's data types [\textbf{here}](https://mc-stan.org/docs/reference-manual/overview-of-data-types.html), 
you should be able to adapt the previous code .stan and .cpp code chunks to 
include more inputs. In order to produce the templated code for these updated 
functions, you should follow the final bullet in the "External binomial hpp code"
subsection. 

Note: This is slightly more difficult in the case of matrices and vectors and these 
types are covered in the "Matrices and vertices" subsection below. Constrained 
data types were also not utilized in this walkthrough. 

`r left_subsection_header("Including more files")`

Depending on the complexity of your external C++ code, it's likely that you 
will need to include more than one C++ file within your CmdStan compilation. 
Luckily, the external C++ file that is supplied to CmdStan doesn't have to be a 
header file (.hpp extension), but it can also be a source code file (.cpp). 
That means your external C++ file can follow this fairly common coding structure: 

- Have a main.cpp file that is called during CmdStan compilation. 
- Within that main.cpp file, you can include the headers for additional .hpp files. 
- Within those .hpp files, you can include their associated .cpp files. 

In this way, when CmdStan attempts to compile, the compiler will first look for 
main.cpp, then the .hpp files that main.cpp points to, and then the .cpp files 
that are pointed to by the .hpp files. 

For example, instead of using the following `make` call: 

```
make STANCFLAGS=--allow-undefined USER_HEADER=examples/binomial_glm/external_binomial_glm.hpp \ 
  examples/binomial_glm/binomial_glm
```

... a user can use the following `make` call instead: 

```
make STANCFLAGS=--allow-undefined USER_HEADER=examples/binomial_glm/external_binomial_glm.cpp \ 
  examples/binomial_glm/binomial_glm
```

Within "external_binomial_glm.cpp" you could have an include statement at the 
top of the file that would link an additional header file. Like so: 

```
#include "external_binomial_glm.hpp"
```

For more information on source file inclusion, refer to this 
[\textbf{reference}](https://en.cppreference.com/w/cpp/preprocessor/include).

`r left_subsection_header("Matrices and vertices")`

When it comes to matrix and vector multiplication, it is more difficult due to 
the fact that Stan uses the 
[\textbf{Eigen library}](https://eigen.tuxfamily.org/dox/group__TutorialMatrixClass.html) 
to represent matrices and the 
[\textbf{standard C++ library}](https://cplusplus.com/reference/vector/vector/)
to represent vectors. As such, your external C++ code will also need to use 
these libraries. 

The difficulty comes with simple operations such as matrix multiplication, which 
are handled within the Eigen library. The Eigen functions would expect certain data 
types within the matrices. However, since those Eigen matrices contain 
Stan data types, you can't use the functions from the Eigen library to handle 
the simple operations because compilation will fail. The C++ vectors 
behave in the same way. 

The current work around to this issue is performing the matrix and vector 
operations within the Stan model code, and then storing the result into a vector 
within the Stan model code. The values can be iterated over afterwards, passed 
into an external function, and then the result is returned. This what was done for 
the binomial GLM example. 

The Stan developers most likely have a solution to this issue. This most likely
be in the form of Stan math functions that take the place of Eigen/C++ functions.
Reaching out to them via the Stan discourse forum would be beneficial in figuring 
out this issue. 

`r left_subsection_header("Limitations")`

The "limit" to CmdStan's external C++ functionality is whether the custom function 
can be auto-differentiated by Stan. If not, then derivatives would need to 
be computed within the custom function. 

However, aside from this unique limit, the typical Stan and C++ limitations 
still take exist. If your Stan model is not syntactically correct, then compilation 
will fail. If you C++ code is not syntactically correct, then compilation will fail. 

It's for these reasons that it might be easier to initially write code for the 
Stan model and C++ code separate from each other. Once both the Stan model and 
C++ code have been testing, the process for templating the C++ can be begin. 

\newpage

`r left_section_header("Aknowledgements")`

[\textbf{Dr. Jay Barber}](https://directory.nau.edu/person/jjb485), the faculty member that oversaw this section of graduate research. 

The Stan team for their [\textbf{extensive documentation}](https://mc-stan.org/users/documentation/) 
and help via the [\textbf{Stan discourse forums}](https://discourse.mc-stan.org/). 

[\textbf{Julian Faraway}](https://julianfaraway.github.io/) since the head injury data that used was sourced from their book, Linear Models with R. 